Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}

### 2019-06-12 doc/paper.tex
adiantado artigo, backup

\begin{document}
%
\title{Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri and M. C. F. de Oliveira}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, Brazil\\
\email{renato.fabbri@gmail.com},
\email{cristina@icmc.usp.br}}
%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
Most real-world networks are dynamic, i.e., they change over time as nodes and links are added or removed, and their representation may also involve variations in the associated data.
There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
a number of procedures have been proposed to visualize them.
  This article presents a visualization method for dynamic scale-free networks.
  Compliant methods were not found in other tools by the authors, which motivated this description and the development of a dedicated interface.
  Named Versinus, the visualization is specially useful in the observation of (in)stability of basic topological characteristics, and consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
  Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
  The Versinus visualization can provide insight into network properties and is useful for observing its behavior, as described in this paper, after a formal presentation of the method and its associated software implementations.
\keywords{Network visualization
12

\and Dynamic networks
\and Longitudinal networks
\and Time-evolving networks
\and Evolving networks
\and Animated visualization
\and Complex networks
\and Data visualization.}
\end{abstract}
%
%
%
\section{Introduction}
The visualization of dynamic networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,hajij,evo3,evo2}.
In fact, the meaningful visualization of dynamic networks is a well-known problem in Information Visualization and dynamic network analysis.
In this work, we present an approach for dynamic network visualization that consists in distributing the nodes along a fixed layout arranged in three segments associated with the three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}. The network dynamic features are expressed by means of changes in visual properties of the glyphs used to depict nodes and links, e.g., glyph height, width, color and shape, and line (arrow) disposition, thickness and color. Moreover, complementary global and local information is rendered in order to enable an acoustic depiction of features of interest.
The method has been called Versinus since its early usages in 2013, because the layout stabilized, after attempts with distinct arrangements, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line). 
There are four software implementations of Versinus, two of them written in Python scripts and packages that yield animations through stop-motion. The other two are written in JavaScript, one focused on exploitation of static networks, the other is an experimental Visual Analytics interface, which extends the visual metaphor to the 3D space and provides additional analysis techniques such as principal component analysis (PCA) and resources for structure navigation\footnote{An instance of the interface is available at: http://rfabbri.vicg.icmc.usp.br:3000/evolution}.
The writing of this document was motivated by
the fact that no method was found by the authors that aims at visualizing dynamic scale-free networks through node-link diagrams.
Furthermore, the visualization method was employed in scientific research~\cite{thesis,stab}, and thus a careful mathematical and technological description of the procedures involved in applying Versinus to real networks is appropriate.

@@

With such structures at hand, the nodes are placed along three segments, as depicted in Figure~\ref{fig:glob}.
The hubs are placed along the first half of a sine period.
The intermediary nodes are placed along the second half of the sine period.
The peripheral nodes are placed along an independent line.
The placement follows 
the basic parameters:
\begin{itemize}
  \item the amplitude $\alpha$ of the sine oscillation.
  \item The displacement $\Delta$ of the sine in the vertical axis.
  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
  \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
\end{itemize}

@@

\subsection{Drawing nodes, links and auxiliary information}
Additional concepts and quantities need to be defined for visualizing the evolving network as a data-driven animation.
A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary message.
The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated at every $\eta$ transactions. To achieve the final animation, a transaction rate $\omega$ must also be defined.
Thus, the following parameters are specified:
\begin{itemize}
  \item the window $\sigma$ captured at each snapshot, expressed in transactions.
  \item the step size $\eta$ between each snapshot, expressed in transactions.
  \item the transaction rate $\omega$, expressed in transactions per second.
\end{itemize}

@@

In each snapshot, the entailed network is analyzed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength, if preferred) and clustering coefficient. Then the visual mapping of information is performed as follows:
\begin{itemize}
  \item nodes not incident in the snapshot are not shown. Glyphs associated with nodes incident in the snapshot are assigned a minimum width and height.
  \item The node sector in the snapshot is mapped to a glyph shape (e.g. circle, hexagon and diamond for hub, intermediary and peripheral nodes, respectively).
  \item Node width is incremented proportionally to its in-degree (or out-strength, if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preferred).
  \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
  \item Link weight is mapped to line width and/or a color scale.
\end{itemize}

@@

\subsection{Network sonification}
The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
Results are of limited effectiveness in the sense that the sonification does not enable users to grasp network features with ease,
specially if they are not musically trained.
Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network),
and is thus probably a unique multimedia representation of networks.
Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented,
i.e. in its final version,
even though a full understanding of the summary requires knowledge of musical theory.
Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated each two tempos.
The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is also updated every two tempos.
The third hub connectivity is proportional to the lower note,
which is incident once a beat, and has its note value updated accordingly.
The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated every note, and is only heard on the second half-beat of every tempo.
Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).

\subsection{Animations produced and available}\label{sec:vid}
Many animations were obtained through the routines described.
These encompassed different networks and settings ($\alpha$, $\sigma$, $\eta$,
etc), and different flavors of Versinus.
Some of these animations are available in six playlists of Youtube videos.
For organization and brevity, they are available at~\cite{verRepo} together with other Versinus-related resources.

@@

The visualization of network structures is favored by the placement of nodes in an oscillatory pattern, as it tends to decrease the superposition of nodes and links. In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
Using a single period for the sine is an arbitrary choice, established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus. Nonetheless, it is allowed to vary in more recent implementations in JavaScript listed in~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
Another important insight is the pertinence of fixed positions for the observation of evolving networks.
Not only this has the potential to reduce the hairball effect~\cite{hairball},
but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images while the network evolves, as in force-based layouts.
The placement of nodes with respect to their topological features (nodes are ordered by connectivity,
segments are related to network sectors) revealed very useful for understanding network features,
such as stability and connectivity-dependent behaviors.
Visual cues such as glyph width, height, shape and color were found informative for grasping information on individual nodes,
but results in a cognitive overload when attempting to observe global features,
which motivates simplified settings.
The sonification of the evolving network added to the exotic flavor of Versinus and made it more attractive.
However, it seems that more informative sonifications should require dedicated research,
training of already musically competent users, and very modest goals. 
Furthermore, reasonable parametrization (e.g. through $\alpha$, $\Delta$, $\mu$, $(x_0, y_0)$, $(x_1, y_1)$, $\sigma$, $\eta$ and $\omega$) is highly dependent on the network characteristics and user intended tasks, and calls for interactive interfaces and automated recommendation of initial values.

>>>>

\begin{document}
%
\title{Versinus: an Animated Visualization Method for Dynamic Scale-Free Networks\thanks{Supported by FAPESP, project 2017/05838-3}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Renato Fabbri\orcidID{0000-0002-9699-629X} \and
Maria Cristina Ferreira de Oliveira\orcidID{0000-0002-4729-5104}}
%
\authorrunning{R. Fabbri and M. C. F. de Oliveira}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{University of São Paulo, São Carlos SP, Brazil\\
\email{renato.fabbri@gmail.com},
\email{cristina@icmc.usp.br}}
%\url{http://conteudo.icmc.usp.br/pessoas/cristina/}}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
% The abstract should briefly summarize the contents of the paper in
% 150--250 words.
Most real-world networks are dynamic, i.e., they change over time as nodes and links are added or removed, and their representation may also involve variations in the associated data.
There is a growing interest in modeling dynamic (or time-evolving, or longitudinal) networks and
a number of procedures have been proposed to visualize them.
  This article presents a visualization method for dynamic scale-free networks.
  Compliant methods were not found in other tools by the authors, which motivated this description and the development of a dedicated interface.
  Named Versinus, the visualization is specially useful in the observation of (in)stability of basic topological characteristics, and consists essentially in placing the most connected nodes (hubs and intermediary) along a sine curve, and the peripheral nodes along a separate segment, usually a line.
  Thus its name, from  Latin \emph{versus} (line) and \emph{sinus} (sine).
  The Versinus visualization can provide insight into network properties and is useful for observing its behavior, as described in this paper, after a formal presentation of the method and its associated software implementations.
\keywords{Network visualization
\and Dynamic networks
\and Longitudinal networks
\and Time-evolving networks
\and Evolving networks
\and Animated visualization
\and Complex networks
\and Data visualization.}
\end{abstract}
%
%
%
\section{Introduction}
The visualization of dynamic networks poses challenges related to data size and complexity which have merited various contributions~\cite{evo1,hajij,evo3,evo2}.
In fact, the meaningful visualization of dynamic networks is a well-known problem in Information Visualization and dynamic network analysis.
In this work, we present an approach for dynamic network visualization that consists in distributing the nodes along a fixed layout arranged in three segments associated with the three sectors (or partitions) of a scale-free network, namely for the hub, intermediary and peripheral nodes~\cite{stab}. The network dynamic features are expressed by means of changes in visual properties of the glyphs used to depict nodes and links, e.g., glyph height, width, color and shape, and line (arrow) disposition, thickness and color. Moreover, complementary global and local information is rendered in order to enable an acoustic depiction of features of interest.
The method has been called Versinus since its early usages in 2013, because the layout stabilized, after attempts with distinct arrangements, in a sine period and a line, which in Latin is \emph{sinus} (sine) and \emph{versus} (line). 
There are four software implementations of Versinus, two of them written in Python scripts and packages that yield animations through stop-motion. The other two are written in JavaScript, one focused on exploitation of static networks, the other is an experimental Visual Analytics interface, which extends the visual metaphor to the 3D space and provides additional analysis techniques such as principal component analysis (PCA) and resources for structure navigation\footnote{An instance of the interface is available at: http://rfabbri.vicg.icmc.usp.br:3000/evolution}.
The writing of this document was motivated by
the fact that no method was found by the authors that aims at visualizing dynamic scale-free networks through node-link diagrams.
Furthermore, the visualization method was employed in scientific research~\cite{thesis,stab}, and thus a careful mathematical and technological description of the procedures involved in applying Versinus to real networks is appropriate.

@@

With such structures at hand, the nodes are placed along three segments, as depicted in Figure~\ref{fig:glob}.
The hubs are placed along the first half of a sine period.
The intermediary nodes are placed along the second half of the sine period.
The peripheral nodes are placed along an independent line.
The placement follows 
the basic parameters:
\begin{itemize}
  \item The amplitude $\alpha$ of the sine oscillation.
  \item The displacement $\Delta$ of the sine in the vertical axis.
  \item The horizontal margin $\mu$, i.e. the distance from the left border to the first hubs, assumed the same as the distance from the right border to last intermediary node.
  \item The positions $(x_0,y_0)$ and $(x_1,y_1)$, i.e. the endpoints of the segment for the peripheral sector.
\end{itemize}

@@

\subsection{Drawing nodes, links and auxiliary information}
Additional concepts and quantities need to be defined for visualizing the evolving network as a data-driven animation.
A \emph{snapshot} $S$ consists of a number $\sigma$ time-contiguous transactions, e.g. 100 email messages, starting at an arbitrary message.
The separation $\eta$ between each snapshot is the number of transactions skipped from one snapshot to the next.
In summary, the evolving network is visualized through a sliding window of $\sigma$ transactions, updated at every $\eta$ transactions. To achieve the final animation, a transaction rate $\omega$ must also be defined.
Thus, the following parameters are specified:
\begin{itemize}
  \item The window $\sigma$ captured at each snapshot, expressed in transactions.
  \item The step size $\eta$ between each snapshot, expressed in transactions.
  \item The transaction rate $\omega$, expressed in transactions per second.
\end{itemize}

@@

In each snapshot, the entailed network is analyzed in terms of the sectors (hubs, intermediary and periphery), in and out degree (or strength, if preferred) and clustering coefficient. Then the visual mapping of information is performed as follows:
\begin{itemize}
  \item Nodes not incident in the snapshot are not shown. Glyphs associated with nodes incident in the snapshot are assigned a minimum width and height.
  \item The node sector in the snapshot is mapped to a glyph shape (e.g. circle, hexagon and diamond for hub, intermediary and peripheral nodes, respectively).
  \item Node width is incremented proportionally to its in-degree (or out-strength, if preferred). Node height is incremented proportionally to its out-degree (or out-strength if preferred).
  \item Node color is related to clustering coefficient, e.g. zero is mapped to white, one is pure red, and intermediary values are scaled accordingly.
  \item Link weight is mapped to line width and/or a color scale.
\end{itemize}

@@

\subsection{Network sonification}\label{sson}
The implementation illustrated in Figure~\ref{fig:final} (b) includes a sonification routine for the evolving network.
Results are of limited effectiveness in the sense that the sonification does not enable users to grasp network features with ease,
specially if they are not musically trained.
Nevertheless, it made the (audio)visualization more engaging, as it yields exotic media,
and it is the first time the authors appreciated a sonification of an evolving network (maybe even of any network),
and is thus probably a unique multimedia representation of networks.
Accordingly, we understand suitable to provide a short description of the sonification as it was left implemented,
i.e. in its final version,
even though a full understanding of the summary requires knowledge of musical theory.
Four sonic/musical voices are used, two of them are noises and the other two have defined pitch. The lower noise has central frequency proportional to first hub connectivity,
which is the node positioned at the extreme left on the layout, on the starting point of the sine, and its frequency is updated every two beats.
The higher noise has central frequency proportional to the second hub, the node next to the first hub, and its frequency is updated at each beat.
The third hub connectivity is proportional to the lower note,
which is incident once a beat, and has its note value updated at every beat.
The fourth hub connectivity is proportional to the higher note, which is incident twice every half-beat, is updated at every note, and is only heard on the second half-beat of every tempo.
Figure~\ref{fson} illustrates such disposition of sonic lines (or musical voices).

\begin{figure}[!h]\centering
\includegraphics[width=.9\textwidth]{sonif}
  \caption{Visual representation of the sonification implemented for the dynamic networks.
  $c_f$ and $f_0$ stands for \emph{center frequency} and \emph{fundamental frequency}, respectively.
  Further details are given in Section~\ref{sson}.
  }\label{fson}
\end{figure}

\subsection{Animations produced and available}\label{sec:vid}
A number of animations were obtained through the routines described.
These encompassed different networks, settings ($\alpha$, $\sigma$, $\eta$,
etc), and different flavors of Versinus.
Some of these visualizations are available in six playlists of Youtube videos.
For organization and brevity, they are available at~\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus} together with other Versinus-related resources.

\subsection{An experimental Visual Analytics interface}
This section exemplifies the employment of the Versinus method
within a more elaborate framework.
Let us consider an interactive visual interface, with which the user may parametrize Versinus and then navigate through the entailed dynamic network.

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis___}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis2___}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis3}
  \caption{
  }\label{esfnetvis}
\end{figure}

\begin{figure}[!h]\centering
\includegraphics[width=\textwidth]{esfNetVis4}
  \caption{
  }\label{esfnetvis}
\end{figure}
One basic feature of the method is the constant representation of
nodes in each of the hubs, intermediary and peripheral sectors.
Nodes are positioned according to such classification regarding all
the activity of interest, and in each snapshot visual cues map the
classification within the snapshot.
Following the analytic methods used in~\cite{stab}, it may be crucial
to know how the relative sizes of the sectors vary along time, i.e. along
the sequence of snapshots.

@@

The visualization of network structures is favored by the placement of nodes in an oscillatory pattern, as it tends to decrease the superposition of nodes and links. In fact, this principle was implemented in some of the layouts available within ccNetViz, and it may be useful in other well-known layouts, such as the Hive Plot~\cite{hive}.
Using a single period for the sine is an arbitrary choice, established because it entailed good visualizations for the networks and tasks at hand during the design of Versinus. Nonetheless, it is allowed to vary in more recent implementations in JavaScript\footnote{Pivot repository for Versinus resources such as implementations, public instances of interactive interfaces, and videos: https://github.com/ttm/versinus}: the sine in which the hub and intermediary nodes are positioned may have as many periods as parametrized by the analyst, and the same is feature is available for the segment for the peripheral nodes.
Another important insight is the pertinence of fixed positions for the observation of evolving networks.
Not only this has the potential to reduce the hairball effect~\cite{hairball},
but it is most often very difficult to keep track of individual nodes and structures when the nodes are allowed to move in the images while the network evolves, as in force-based layouts.
The placement of nodes with respect to their topological features (nodes are ordered by connectivity,
segments are related to network sectors) revealed very useful for understanding network features,
such as stability and connectivity-dependent behaviors.
Visual cues such as glyph width, height, shape and color were found informative for grasping information on individual nodes,
but results in a cognitive overload when attempting to observe global features,
which motivates simplified settings.
The sonification of the evolving network added to the exotic flavor of Versinus and made it more attractive.
However, it seems that more informative sonifications should require dedicated research,
training of already musically competent users, and very modest goals. 
Furthermore, reasonable parametrization (e.g. through $\alpha$, $\Delta$, $\mu$, $(x_0, y_0)$, $(x_1, y_1)$, $\sigma$, $\eta$ and $\omega$) is highly dependent on the network characteristics and user intended tasks, and calls for interactive interfaces and automated recommendation of initial values.
